### 2014-08-22 src/docs/talks/msr-nyc-2014/stan-current-future.tex
version of talk for presentation

\mypart{{\LARGE The Present:} \\ \spc\spc{\Huge Stan 2.4}}

@@

\begin{itemize}
\item Stan is an \myemph{imperative} probabilistic programming language
\vspace*{-12pt}
\begin{itemize}\footnotesize
\item  cf., BUGS: declarative; \ Church: functional; \ Figaro: object-oriented
\end{itemize}
\item Stan \myemph{program}
\vspace*{-4pt}
\begin{itemize}\small
\item declares variables
\item codes log posterior (penalized likelihood)
\item calculates gradients, Hessians, etc., using auto-diff
\end{itemize}
\item Stan \myemph{inference}
\vspace*{-4pt}
\begin{itemize}\small
\item MCMC for full Bayesian inference (+ posterior analysis)
\item MLE for point estimates (+ Hessian-based error estimates)
\end{itemize}
\end{itemize}

@@

\sld{Who's Using Stan?}
\begin{itemize}
\item 830+ registrations for mailing list
\item spanning physical, biomedical, and social sciences, 
\\ plus engineering, education, finance, and marketing

\item Papers citing Stan on {\slshape Google Scholar} in subjects
\\[3pt]
{\footnotesize clinical drug
trials, general computational statistics, entomology, opthalmology,
neurology, sociology and population dynamics, genomics, agriculture,
psycholinguistics, molecular biology, population dynamics, materials
engineering, botany, astrophysics, oceanography, election prediction,
fisheries, cancer biology, public health and epidemiology, population
ecology, collaborative filtering for recommender systems, climatology,
educational testing, natural language processing, econometrics}
\end{itemize}

\sld{Documentation and Examples}
%
\begin{itemize}
\item 400+ page core language tutorial manual
\item separate installation, getting started, interface manuals 
\\ {\footnotesize  (RStan, PyStan, CmdStan)}
\item BUGS and JAGS examples (all 3 volumes), 
\item Gelman and Hill, {\slshape Data Analysis Using Regression and
    Multilevel/Hierarchical Models}
\item Wagenmakers and Lee, {\slshape Bayesian Cognitive Modeling}
\item two other books in progress
\item user-contributed examples in group
\end{itemize}

\sld{Scaling and Evaluation}
\begin{itemize}
\item Type of scaling \hfill (improvement$^{*}$)
\vspace*{-4pt}
\begin{itemize}\small
\item more data \hfill (1--2 orders of magnitude)
\item more parameters \hfill (1--2 orders of magnitude)
\item \myemph{more complex models} \hfill ({\large \myemph{$\infty$}})
\end{itemize}
\item imporvement vs.\ ``state-of-the-art'' \myemph{black-box MCMC}
measured by
\vspace*{-4pt}
\begin{itemize}\small
\item time to convergence
\item time per effective sample at mixing
\item memory usage
\end{itemize}
\end{itemize}

\mypart{Part I}{Stan Front End}

@@

\begin{itemize}
\item \myemph{\tt\bfseries data} \ (once) 
\vspace*{-4pt}
\begin{itemize}\small
\item {\slshape content}: declare data types, sizes, and constraints
\item {\slshape execute}: read from data source, validate constraints
\end{itemize}
%
\item \myemph{\tt\bfseries parameters} \ (every log prob eval)
\vspace*{-4pt}
\begin{itemize}\small
\item {\slshape content}: declare parameter types, sizes, and constraints
\item {\slshape execute}: transform to constrained, optional Jacobian
\end{itemize}
%
\item \myemph{\tt\bfseries model} \ (every log prob eval) 
\vspace*{-4pt}
\begin{itemize}\small
\item {\slshape content}: statements definining posterior density
\item {\slshape execute}: execute statements
\end{itemize}
\end{itemize}

@@

\begin{itemize}
\item All built-in \myemph{C++ functions and operators}
\\
{\footnotesize C math, TR1, C++11, including all trig, pow, and
  special log1m, erf, erfc, fma, atan2, etc.}
\item Extensive library of \myemph{statistical functions}
\\
{\footnotesize e.g., softmax,
log gamma and digamma functions, beta functions, Bessel functions of
 and second kind, etc.}
%
\item Efficient, arithmetically stable \myemph{compound functions}
\\
{\footnotesize e.g., multiply log, log sum of
exponentials, log inverse logit}
\end{itemize}

@@

\begin{itemize}
\item \myemph{Basic arithmetic}: all arithmetic operators
\item \myemph{Elementwise arithmetic}: vectorized operations
\item \myemph{Solvers}: matrix division, (log) determinant,
inverse 
\item \myemph{Decompositions}: QR, Eigenvalues and Eigenvectors, 
\\
Cholesky factorization, singular value decomposition
\item \myemph{Compound Operations}: quadratic forms, variance scaling
\item \myemph{Ordering, Slicing, Broadcasting}: sort, rank, block, rep
\item \myemph{Reductions}: sum, product, norms
\item \myemph{Specializations}: triangular, positive-definite, etc.
\end{itemize}

@@

\begin{itemize}
\item Each distribution has
\vspace*{-4pt}
\begin{itemize}\small
\item log density or mass function
\item cumulative distribution functions, plus complementary versions,
  plus log scale
\item pseudo Random number generators
\end{itemize}
\item Alternative parameterizations
\\
{\footnotesize (e.g., Cholesky-based multi-normal,
log-scale Poisson, logit-scale Bernoulli)}
\item New multivariate correlation matrix density: LKJ
\\
{\footnotesize degrees of freedom controls 
shrinkage to (expansion from) unit matrix}
\end{itemize}

\sld{Statements}

\begin{itemize}
\item \myemph{Sampling}: \ {\footnotesize \Verb|y ~ normal(mu,sigma)|}
\ \ \ {\footnotesize (increments log probability)}
%
\item \myemph{Log probability}: \ {\footnotesize increment\_log\_prob(lp);}
\item \myemph{Assignment}: \  {\footnotesize \code{y\_hat <- x * beta};}
%
\item \myemph{For loop}: \ {\footnotesize \code{for (n in 1:N) ...}}
%
\item \myemph{While loop}: \ {\footnotesize \code{while (cond) ...}}
%
\item \myemph{Conditional}: \ {\footnotesize
\code{if (cond) ...; else if (cond) ...;  else ...;}}
\item \myemph{Block}: \ {\footnotesize \Verb|{ ... }|}  \ \ \ {\footnotesize
   (allows local variables)}
\item \myemph{Print}: \ {\footnotesize print("theta=",theta);}
\end{itemize}

@@

\begin{itemize}
\item Generated quantities block for \myemph{inference}
\\ {\footnotesize (predictions, decisions, and event probabilities)}
\item \myemph{Extractors} for samples in RStan and PyStan
\item Coda-like \myemph{posterior summary}
\vspace*{-4pt}
\begin{itemize}\small
\item posterior mean w.\ standard error, standard deviation, quantiles
\item split-$\hat{R}$ multi-chain convergence diagnostic (Gelman and ...)
\item multi-chain effective sample size estimation (FFT algorithm)
\end{itemize}
\item Model comparison with \myemph{WAIC}
\\
{\footnotesize (internal log likelihoods; external cross-sample statistics)}
\end{itemize}

@@

\mypart{Part II}{Under the Hood}

@@

\mypart{{\LARGE The Future:} \\ \spc\spc{\Huge Stan 3 \& Beyond}}

\sld{Differential Equation Solver}
\begin{itemize}
\item Auto-diff solutions w.r.t.\ parameters
\item Integrate coupled system for solution with partials
\item Auto-diff coupled Jacobian for stiff systems
\vfill
\item C++ prototype integrated for large PK/PD models
\vspace*{-4pt}
\begin{itemize}\footnotesize
\item 
Project with Novartis: 
  longitudinal clinical trial w.\ multiple drugs, dosings, placebo control,
  hierarchical model of patient-level effects, meta-analysis
\end{itemize}
\vfill
\item Generalized code complete, under testing
\item {\footnotesize With: Frederic Bois, Amy Racine-Poon, Sebastian Weber}
\end{itemize}

\sld{MLE/MML Error Estimation}
\begin{itemize}
\item Use curvature to estimate error on unconstrained scale
\item Transform back to constrained scale 
\item May require sampling from Laplace approximation
\vfill
\item Design complete
\end{itemize}

\mypart{Appendix I}{Bayesian Data Analysis}

>>>>

\mypart{Current:}{Stan 2.4}

@@

\begin{itemize}
\item Stan is an \myemph{imperative} probabilistic programming language
\vspace*{-12pt}
\begin{itemize}\footnotesize
\item  cf., BUGS: declarative; \ Church: functional; \ Figaro: object-oriented
\end{itemize}
\item Stan \myemph{program}
\vspace*{-4pt}
\begin{itemize}\small
\item declares variables
\item codes log posterior (or penalized likelihood)
\end{itemize}
\item Stan \myemph{inference}
\vspace*{-4pt}
\begin{itemize}\small
\item Sampling for full Bayesian inference
\item Optimization + curvature for maximum likelihood estimates
\end{itemize}
\end{itemize}

@@

\sld{Who's Using Stan?}
\begin{itemize}
\item 830+ user mailing list; 100+ citations
\vspace*{-4pt}
\begin{itemize}\small
\item physical, biomedical, and social sciences
\item plus engineering, education, finance, and marketing
\end{itemize}
\item Application areas
  \\[3pt]
  {\footnotesize clinical drug trials, general computational
    statistics, entomology, opthalmology, neurology, sociology and
    population dynamics, genomics, agriculture, psycholinguistics,
    molecular biology, population dynamics, materials engineering,
    botany, astrophysics, oceanography, election prediction,
    fisheries, cancer biology, public health and epidemiology,
    population ecology, collaborative filtering, climatology,
    educational testing, natural language processing, econometrics}
\end{itemize}

\sld{Documentation and Examples}
%
\begin{itemize}
\item Documentation
\vspace*{-4pt}
\begin{itemize}\footnotesize
\item 400+ page user's guide and reference
\item For each interface: installation, getting started, reference
\end{itemize}
\item Examples
\vspace*{-4pt}
\begin{itemize}\footnotesize
\item BUGS and JAGS examples (all 3 volumes), 
\item Gelman and Hill, {\slshape Data Analysis Using Regression and \\
    Multilevel/Hierarchical Models}
\item Wagenmakers and Lee, {\slshape Bayesian Cognitive Modeling}
\item two books in progress
\item user-contributed examples in group
\end{itemize}
\end{itemize}

\sld{Scaling and Evaluation}
\begin{itemize}
\item Type of scaling
\vspace*{-4pt}
\begin{itemize}\small
\item more data (e.g., observations)
\item more parameters (e.g., regression coefficients)
\item \myemph{more complex models} (e.g., multilevel priors)
\end{itemize}
\item Metrics
\vspace*{-4pt}
\begin{itemize}\small
\item time to convergence
\item time per effective sample after convergence
\item memory usage
\vfill
\item 0 to {\large $\infty$} orders of magnitude improvement on
  black-box state-of-the-art (e.g., BUGS, JAGS)
\item more improvement with more complex models
\end{itemize}
\end{itemize}

@@

\begin{itemize}
\item \myemph{\tt\bfseries data} \ (once) 
\vspace*{-4pt}
\begin{itemize}\small
\item {\slshape content}: declare data types, sizes, and constraints
\item {\slshape execute}: read from data source, validate constraints
\end{itemize}
%
\item \myemph{\tt\bfseries parameters} \ (every log prob eval)
\vspace*{-4pt}
\begin{itemize}\small
\item {\slshape content}: declare parameter types, sizes, and constraints
\item {\slshape execute}: transform to constrained, optional Jacobian
\end{itemize}
%
\item \myemph{\tt\bfseries model} \ (every log prob eval) 
\vspace*{-4pt}
\begin{itemize}\small
\item {\slshape content}: definine posterior density (up
  to constant)
\item {\slshape execute}: execute statements
\end{itemize}
\end{itemize}

@@

\begin{itemize}
\item All built-in \myemph{C++ functions and operators}
\\
{\footnotesize C math, TR1, C++11, including all trig, pow, and
  special log1m, erf, erfc, fma, atan2, etc.}
\item Extensive library of \myemph{statistical functions}
\\
{\footnotesize e.g., softmax,
log gamma and digamma functions, beta functions, Bessel functions of
first and second kind, etc.}
%
\item Efficient, arithmetically stable \myemph{compound functions}
\\
{\footnotesize e.g., multiply log, log sum of
exponentials, log inverse logit}
\end{itemize}

@@

\begin{itemize}
\item \myemph{Type inference}, {\small e.g., row vector times vector is scalar}
\item \myemph{Basic and elementwise arithmetic}: all ops
\item \myemph{Solvers}: matrix division, (log) determinant,
inverse 
\item \myemph{Decompositions}: QR, Eigenvalues and Eigenvectors, 
\\
Cholesky factorization, singular value decomposition
\item \myemph{Ordering, Slicing, Broadcasting}: sort, rank, block, rep
\item \myemph{Reductions}: sum, product, norms
\item \myemph{Compound Operations}: quadratic forms, variance scaling
\item \myemph{Specializations}: triangular, positive-definite, etc.
\end{itemize}

@@

\begin{itemize}
\item Each distribution has
\vspace*{-4pt}
\begin{itemize}\small
\item log density or mass function
\item cumulative distribution functions, plus complementary versions,
  plus log scale
\item pseudo Random number generators
\end{itemize}
\item Alternative parameterizations
\\
{\footnotesize (e.g., Cholesky-based multi-normal,
log-scale Poisson, logit-scale Bernoulli)}
\item New multivariate correlation matrix density: LKJ
\\
{\footnotesize degrees of freedom controls 
shrinkage to (expansion from) unit matrix; \\
independently scale by parameter}
\end{itemize}

\sld{Statements}
\vspace*{-2pt}
\begin{itemize}
\item \myemph{Sampling}: \ {\footnotesize \Verb|y ~ normal(mu,sigma)|}
\ \ \ {\footnotesize (increments log probability)}
%
\item \myemph{Log probability}: \ {\footnotesize increment\_log\_prob(lp);}
\item \myemph{Assignment}: \  {\footnotesize \code{y\_hat <- x * beta};}
%
\item \myemph{For loop}: \ {\footnotesize \code{for (n in 1:N) ...}}
%
\item \myemph{While loop}: \ {\footnotesize \code{while (cond) ...}}
%
\item \myemph{Conditional}: \ {\footnotesize
\code{if (cond) ...; else if (cond) ...;  else ...;}}
\item \myemph{Block}: \ {\footnotesize \Verb|{ ... }|}  \ \ \ {\footnotesize
   (allows local variables)}
\item \myemph{Print}: \ {\footnotesize \code{print("theta=",theta);}}
\item \myemph{Throw}: \ {\footnotesize \code{raise\_exception("x must be
    positive; found x=", x);}}
\end{itemize}

@@

\begin{itemize}
\item Generated quantities block for \myemph{inference}
\\ {\footnotesize (predictions, decisions, and event probabilities)}
\item \myemph{Extractors} for samples in RStan and PyStan
\item Coda-like \myemph{posterior summary}
\vspace*{-4pt}
\begin{itemize}\small
\item posterior mean w.\ standard error, standard deviation, quantiles
\item split-$\hat{R}$ multi-chain convergence diagnostic (Gelman and ...)
\item multi-chain effective sample size estimation (FFT algorithm)
\end{itemize}
\item Model comparison with \myemph{WAIC} or cross-validation
\\
{\footnotesize (internal log likelihoods; external cross-sample statistics)}
\end{itemize}

@@

\mypart{Future:}{Stan 3 \& Beyond}

\sld{Differential Equation Solver}
\begin{itemize}
\item Auto-diff solutions w.r.t.\ parameters \& initial states
\item Integrate coupled system for solution with partials
\item Auto-diff coupled Jacobian for stiff systems
\vfill
\item C++ prototype integrated for large PK/PD models
\vspace*{-4pt}
\begin{itemize}\footnotesize
\item 
Project with Novartis: 
  longitudinal clinical trial w.\ multiple drugs, dosings, placebo control,
  hierarchical model of patient-level effects, meta-analysis
\end{itemize}
\vfill
\item Generalized code complete, under testing
\item {\footnotesize With: Frederic Bois, Amy Racine-Poon, Sebastian Weber}
\end{itemize}

\sld{Ensemble MCMC}
\begin{itemize}
\item Markov chain is ensemble of parameter vectors
\item Interpolate/extrapolate in ensemble for Metropolis proposal
\begin{itemize}\small
\item Walk \& stretch moves (Goodman \& Weare)
\item Differential evolution (ter Braak)
\end{itemize}
\item Highly parallelizable
\item Derivative free
\vfill
\item Code complete; awaiting integration
\end{itemize}

\sld{MLE \& MML Errors}
\begin{itemize}
\item Simple posterior approximation
\item Standard errors for estimators
\item Approximate posterior covariance with multivariate normal
\item Estimate with Laplace approximation (based on curvature)
\item Sample and transform to constrained scale
\vfill
\item Design complete
\end{itemize}

\mypart{Appendix I}{Under the Hood}

@@

\mypart{Appendix II}{Bayesian Data Analysis}
